---
title: "Data Cleaning"
output: 
  html_document:
    toc: true
    toc_float: true
    code_folding: hide
    
---

```{r setup, include=FALSE}
library(tidyverse)
library(readxl)
library(countrycode)
```

# Source

Since there are many determinants and indicators of violence, we chose the 
indicators and outcomes that seem most relevant and interesting in 
exploring violence from the following sources:

* United Nations Development Program (UNDP): 
  - Human Development Index
* United Nations Office of Drugs and Crime (UNODC)
  - Corruption and Economic Crime
  - Criminal Justice Personnel
  - Human Trafficking
  - Intentional Homicide
  - Violent and Sexual Crimes
* World Bank
  - Gross Domestic Product (GDP)
  - Inflation Rate (Measured by Consumer Price Index)
  - Unemployment Rate
* World Health Organization (WHO)
  - Alcohol Consumption

Intentional homicide, and violent and sexual crimes were chosen as the outcome 
variable to quantify violence. 

* Intentional homicide:
  - counting unit: number of homicide victims per 100,000 population
  - classification:
    - situational context: organized crime, interpersonal (excluding 
    familial/intimate), socio-political
    - relationship to perpetrator: intimate partner, family member, friend, 
    colleague, etc.
    - mechanism: firearm, weapon, physical force

* Violent and sexual crimes:
  - counting unit: number of individual offences per 100,000 population
  - classification of offenses:
    - rape
    - serious assault
    - kidnapping
    - sexual violence
  - NOTE: some countries used other counting unit (a series of offenses to form 
  a case, or multiple cases to form an investigation)

# Scraping Method

The datasets were downloaded from the official websites of the sources above. 
The names of the files were also changed accordingly for clarity and to avoid 
confusion. For example, the alcohol consumption data file was renamed from 
"data.csv" to "alcohol_consumption.csv".

# EDA Data

For EDA, since we are interested in exploring the trends of global violence, 
we imported and tidied the datasets for our two outcome variables: homicide 
and violence. 

## Homicide
```{r}
homicide_visual_df = 
    read_excel(path = "data/unodc/intentional_homicide.xlsx",
               skip = 2) |>
  janitor::clean_names() |>
  filter(unit_of_measurement == "Rate per 100,000 population",
         indicator == "Victims of intentional homicide",
         category != "Total",
         sex != "Total",
         between(year, 2015, 2023)) |>
  select(country, region, category, sex, year, 
         homicide_rate = value) |>
  mutate(category = str_replace_all(category, 
                                    "^Intimate partner or family member.*", 
                                    "Intimate partner or family member"),
         category = str_replace_all(category, 
                                    "^Socio-political homicide.*", 
                                    "Socio-political homicide"))

```

The tidied homicide dataset includes country, region, category of homicide 
victims, sex, year and victim rate of homicide. The dataset included data for 
both counts and rates per 100,000 people. Therefore, we filtered to choose rates 
per 100,000 people since it allows standardization for comparability by 
adjusting for differences in population size.

## Violence
```{r}
violence_visual_df = 
    read_excel(path = "data/unodc/violent_sexual_crime.xlsx",
               skip = 2) |>
  janitor::clean_names() |>
  filter(unit_of_measurement == "Rate per 100,000 population",
         between(year, 2015, 2023)) |>
  select(country, region, indicator, category, sex, year, 
         violence_rate = value) |>
  mutate(category = str_replace_all(category, 
                                    "^Sexual violence.*", 
                                    "Sexual violence"))
```

The tidied violence dataset includes country, region, indicator, category of 
violence, sex, year and violence rate. The indicator includes different rates 
for violence (violent offences and victims) which will be explored further in 
EDA. Rates per 100,000 people were chosen also for this dataset for 
comparability. 

# Regression Data

For regression, the datasets for all the chosen predictors and outcomes will be 
tidied and merged for modeling process and analysis. 

## Function to Tidy Datasets
```{r pivot_tidy_function}
tidy_df = function(data, variable, pivot = FALSE, average = FALSE) {
  
  if(pivot) {
    data = data |>
      pivot_longer(cols = -iso3_code,
                   names_to = "year",
                   values_to = variable) |>
      janitor::clean_names()
  }
  
  data = data |>
    mutate(year = as.numeric(year),
           iso3_code = str_replace_all(iso3_code, "^GBR.*", "GBR"),
           iso3_code = str_replace_all(iso3_code, "^IRQ.*", "IRQ"),
           country = countrycode(iso3_code, origin = "iso3c",
                                 destination = "country.name",
                                 nomatch = NA, warn = FALSE)) |> 
    drop_na(country)
  
  if(average) {
    
    if("region" %in% colnames(data)) {
      data = data |>
        group_by( country, region, year) |>
        summarize(!!variable := mean(value))
    }
    
    else {
        data = data |>
          group_by(country, year) |>
          summarize(!!variable := mean(value))
    }
    
  }
  
  return(data)
  
}
```

We defined a function `tidy_df()`to pipe after importing each dataset to 
automatically tidy the datasets. The `countrycode()` function standardizes the 
country names of each dataset since some of the country names are represented 
inconsistently across dataset files. For example, South Korea was represented 
as "Korea (Republic of)" and "Korea, Rep." 

Moreover, for the countrycode() function, NA is returned if no matches are 
found and warnings are also suppressed by `nomatch = NA` and `warn = FALSE` 
respectively since some datasets include iso3 codes for entities that are not 
countries. For example, EUU represents the European Union; however, the 
`countrycode()` function only recognizes iso3 codes for countries and returns a 
warning, but we are interested in countries and regions. A boolean parameter 
called `pivot` was included to indicate if a dataset needs to be pivoted and 
subsequently perform `pivot_longer()`, since some datasets had years organized 
as different columns, where each column represented values for that year. 

Also, some of the datasets include rates for different categories. For example, 
the data for economic crime includes crime rates for fraud, burglary and more. 
Therefore, another boolean parameter called `average` was included to pass an 
option to group by and summarize rates across categories and genders of 
different datasets, in order to provide a standardized measure and reduce 
complexity for analysis.

## Economic Determinants
```{r tidy_econ_determinants, message = FALSE}
gdp_df = 
  read_excel(path = "data/worldbank/gdp.xls",
             sheet = "Data", skip = 3, na = "") |>
  select(iso3_code = 2, "2015":"2023") |>
  tidy_df("gdp", pivot = TRUE)

inflation_df = 
  read_excel(path = "data/worldbank/inflation_rate.xls", 
             sheet = "Data", skip = 3, na = "") |>
  select(iso3_code = 2, "2015":"2023") |>
  tidy_df("inflation_rate", pivot = TRUE)

unemployment_df =
  read_excel(path = "data/worldbank/unemployment_rate.xlsx",
             range = "A1:BP437", na = "..") |>
  rename_with(str_replace_all, pattern = " \\[.*", replacement = "") |>
  select(iso3_code = 4, "2013":"2023") |>
  tidy_df("unemployment_rate", pivot = TRUE)

human_develop_df =
  read_csv(file = "data/undp/human_development_index.csv", na = "") |>
  head(-11) |>
  rename_with(str_replace_all, pattern = "hdi_", replacement = "") |>
  select(iso3_code = 1, "2015":"2022") |>
  tidy_df("hdi", pivot = TRUE)
```

The following variables were identified as economic prdictors of violence: 

* `gdp`
* `inflation`
* `unemployment`
* `human development index`

The `pivot` parameter of the `tidy_df()` function was set to true for these 
datasets, since all of them had the different years as data columns, which 
needs to be pivoted accordingly. 

## Social Determinants
```{r import_social_determinants, message = FALSE}
econ_crime_df = 
  read_excel(path = "data/unodc/corruption_economic_crime.xlsx", skip = 2) |>
  janitor::clean_names() |>
  filter(unit_of_measurement == "Rate per 100,000 population") |>
  tidy_df("crime_rate",  average = TRUE)

personnel_df =
  read_excel(path = "data/unodc/criminal_justice_personnel.xlsx", skip = 2) |>
  janitor::clean_names() |>
  filter(unit_of_measurement == "Rate per 100,000 population",
         indicator == "Criminal Justice Personnel", 
         sex == "Total") |>
  tidy_df("personnel_rate", average = TRUE)

trafficking_df =
  read_excel(path = "data/unodc/human_trafficking.xlsx", skip = 2) |>
  janitor::clean_names() |>
  filter(indicator == "Detected trafficking victims",
         category == "Total", 
         sex == "Total",
         age == "Total",
         txt_value != "<5") |>
  mutate(value = str_replace_all(txt_value, ",", ""),
         value = as.numeric(value)) |>
  select(iso3_code : region, year, trafficked_victims = value) |>
  tidy_df("trafficked_victims")

alcohol_df =
  read_csv(file = "data/who/alcohol_consumption.csv", na = "") |>
  janitor::clean_names() |>
  filter(dim1 == "Both sexes") |>
  select(iso3_code = spatial_dim_value_code, 
         year = period, 
         value = fact_value_numeric) |>
  tidy_df("alcohol_consumption_rate", average = TRUE)
```

The following variables were identified as social prdictors of violence: 

* `economic crime`
* `criminal justice personnel`
* `human trafficking`
* `alcohol`

The datasets for economic crimes and criminal justice personnel also provided 
data for both counts and rates per 100,000 people. Subsequently, we filtered to 
choose rates per 100,000 people for standardization and comparability. The 
human trafficking dataset provided data for number of human trafficking victims, 
instead of trafficking rates; the dataset also included a category for total 
number of trafficking victims, so the data was filtered for total category and 
neither the `pivot` nor `average` options were set to true in the `tidy_df()` 
function.

## Outcomes of Violence
```{r import_violence_outcomes, message = FALSE}
homicide_df = 
  read_excel(path = "data/unodc/intentional_homicide.xlsx", skip = 2) |>
  janitor::clean_names() |>
  filter(indicator == "Victims of intentional homicide",
         unit_of_measurement == "Rate per 100,000 population",
         dimension == "Total", 
         sex == "Total", 
         age == "Total") |>
  tidy_df("homicide_rate", average = TRUE)
  
violence_df = 
  read_excel(path = "data/unodc/violent_sexual_crime.xlsx", skip = 2) |>
  janitor::clean_names() |>
  filter(indicator == "Violent offences",
         unit_of_measurement == "Rate per 100,000 population") |>
  tidy_df("violence_rate", average = TRUE)
```

Unlike the tidied dataset for EDA, the datasets for homicide and violence had 
their respective rates averaged across category, since we are interested in 
the average rates of homicide victims and violent offences for regression 
analysis. 

## Merge Datasets
```{r merge_datasets, message = FALSE}
merged_violence_df =
  left_join(homicide_df, violence_df) |>
  left_join(gdp_df) |>
  left_join(inflation_df) |>
  left_join(unemployment_df) |>
  left_join(human_develop_df) |>
  left_join(econ_crime_df) |>
  left_join(personnel_df) |>
  left_join(trafficking_df) |>
  left_join(alcohol_df) |>
  mutate(country = as.factor(country),
         region = as.factor(region)) |>
  filter(between(year, 2015, 2023)) |>
  relocate(iso3_code) |>
  drop_na(violence_rate, homicide_rate)

head(merged_violence_df)
write.csv(merged_violence_df, file = "./data/merged_violence.csv")
```

The final merged dataset includes `r nrow(merged_violence_df)` rows and 
`r ncol(merged_violence_df)` columns, including iso3 codes, country, region, 
year, homicide rate, average violence offence rate, gdp, inflation rate, 
unemployment rate, average crime rate, criminal justice personnel rate, total, 
human trafficking rate and alcohol consumption rate as variables. There is a 
total of `r length(unique(pull(merged_violence_df, country)))` distinct 
countries. 

After merging the datasets, country and region were converted to categorical 
variables and the dataset was filtered for the years between 2015 and 2023, 
the period of our interest for analysis. Although the datasets were merged with 
matching country names tidied by the `countrycode()`, the iso3 codes were 
retained, since they are required for plotting maps which can be found under the 
shiny section of this website. 
