---
title: "P8105 Final Project: Determinants of Global Violence"
author: "My An Huynh, Jeffrey Lin, Soo Min You, Hyun Kim, Malika Top"
output: github_document
---

```{r setup, include=FALSE}
library(tidyverse)
library(readxl)
library(countrycode)
```

## Motivation


(ONE OR TWO PARAGRAPHS)


## Initial Questions


(ONE PARAGRAPH WITH LIST OF QUESTIONS)


# Data: Source, Scraping Method & Cleaning

## Source

Since there are many determinants and indicators of violence, we chose the 
indicators and outcomes that we thought were most interesting and relevant in 
exploring violence from the following sources:

* United Nations Development Program (UNDP): 
  - Human Development Index
* United Nations Office of Drugs and Crime (UNODC)
  - Corruption and Economic Crime
  - Criminal Justice Personnel
  - Human Trafficking
  - Intentional Homicide
  - Violent and Sexual Crimes
* World Bank
  - Gross Domestic Product (GDP)
  - Inflation Rate (Measured by Consumer Price Index)
  - Unemployment Rate
* World Health Organization (WHO)
  - Alcohol Consumption

Intentional homicide, and violent and sexual crimes were chosen as the outcome 
variable to quantify violence. 

* Intentional homicide:
  - counting unit: number of homicide victims per 100,000 population
  - classification:
    - situational context: organized crime, interpersonal (excluding 
    familial/intimate), socio-political
    - relationship to perpetrator: intimate partner, family member, friend, 
    colleague, etc.
    - mechanism: firearm, weapon, physical force

* Violent and sexual crimes:
  - counting unit: number of individual offences per 100,000 population
  - classification of offenses:
    - rape
    - serious assault
    - kidnapping
    - sexual violence
  - NOTE: some countries used other counting unit (a series of offenses to form 
  a case, or multiple cases to form an investigation)

## Scraping Method

The datasets were downloaded from the official websites of the sources above. 
The names of the files were also changed accordingly for clarity and to avoid 
confusion. For example, the alcohol consumption data file was renamed from 
"data.csv" to "alcohol_consumption.csv".

## Cleaning

### Function to Tidy Datasets
```{r pivot_tidy_function}
tidy_df = function(data, variable, pivot = FALSE, average = FALSE) {
  
  if(pivot) {
    data = data |>
      pivot_longer(cols = -iso3_code,
                   names_to = "year",
                   values_to = variable) |>
      janitor::clean_names()
  }
  
  data = data |>
    mutate(year = as.numeric(year),
           iso3_code = str_replace_all(iso3_code, "^GBR.*", "GBR"),
           iso3_code = str_replace_all(iso3_code, "^IRQ.*", "IRQ"),
           country = countrycode(iso3_code, origin = "iso3c",
                                 destination = "country.name",
                                 nomatch = NA, warn = FALSE)) |> drop_na()
  
  if(average) {
    
    if("region" %in% colnames(data)) {
      data = data |>
        group_by(country, region, year) |>
        summarize(!!variable := mean(value))
    }
    
    else {
        data = data |>
          group_by(country, year) |>
          summarize(!!variable := mean(value))
    }
    
  }
  
  return(data)
  
}
```

For the data cleaning process, we defined a function to pipe after importing 
each dataset to automatically tidy the datasets. The countrycode() function 
standardizes the country names of each dataset since some of the country names 
are represented inconsistently across dataset files. For example, South Korea 
was represented as "Korea (Republic of)" and "Korea, Rep." 

Moreover, for the countrycode() function, NA is returned if no matches are 
found and warnings are also suppressed by `nomatch = NA` and `warn = FALSE` 
respectively since some datasets include iso3 codes for entities that are not 
countries. For example, EUU represents the European Union; however, the 
countrycode() function only recognizes iso3 codes for countries and returns a 
warning, but we are interested in countries and regions. A boolean parameter 
called `pivot` was included to indicate if a dataset needs to be pivoted and 
subsequently perform pivot_longer(), since some datasets had years organized as 
different columns, where each column represented values for that year. 

Also, some of the datasets include rates for different categories. 
For example, the data for economic crime includes crime rates for fraud, 
burglary and more. Therefore, another boolean parameter called "average" was 
included to pass an option to group by and summarize rates across categories 
and genders of different datasets, which will then be merged into a single 
dataset to be used for EDA and especially regresion analysis. 

### Economic Determinants
```{r tidy_econ_determinants}
gdp_df = 
  read_excel(path = "data/worldbank/gdp.xls",
             sheet = "Data", skip = 3, na = "") |>
  select(iso3_code = 2, "2015":"2023") |>
  tidy_df("gdp", pivot = TRUE)

inflation_df = 
  read_excel(path = "data/worldbank/inflation_rate.xls", 
             sheet = "Data", skip = 3, na = "") |>
  select(iso3_code = 2, "2015":"2023") |>
  tidy_df("inflation_rate", pivot = TRUE)

unemployment_df =
  read_excel(path = "data/worldbank/unemployment_rate.xlsx",
             range = "A1:BP437", na = "..") |>
  rename_with(str_replace_all, pattern = " \\[.*", replacement = "") |>
  select(iso3_code = 4, "2013":"2023") |>
  tidy_df("unemployment_rate", pivot = TRUE)

human_develop_df =
  read_csv(file = "data/undp/human_development_index.csv", na = "") |>
  head(-11) |>
  rename_with(str_replace_all, pattern = "hdi_", replacement = "") |>
  select(iso3_code = 1, "2015":"2022") |>
  tidy_df("hdi", pivot = TRUE)
```

### Social Determinants
```{r import_social_determinants}
econ_crime_df = 
  read_excel(path = "data/unodc/corruption_economic_crime.xlsx", skip = 2) |>
  janitor::clean_names() |>
  filter(unit_of_measurement == "Rate per 100,000 population") |>
  tidy_df("crime_rate",  average = TRUE)

personnel_df =
  read_excel(path = "data/unodc/criminal_justice_personnel.xlsx", skip = 2) |>
  janitor::clean_names() |>
  filter(unit_of_measurement == "Rate per 100,000 population",
         indicator == "Criminal Justice Personnel", 
         sex == "Total") |>
  tidy_df("personnel_rate", average = TRUE)

trafficking_df =
  read_excel(path = "data/unodc/human_trafficking.xlsx", skip = 2) |>
  janitor::clean_names() |>
  filter(category == "Total", 
         sex == "Total",
         age == "Total",
         txt_value != "<5") |>
  mutate(value = str_replace_all(txt_value, ",", ""),
         value = as.numeric(value)) |>
  tidy_df("trafficking_rate", average = TRUE)

alcohol_df =
  read_csv(file = "data/who/alcohol_consumption.csv", na = "") |>
  janitor::clean_names() |>
  filter(dim1 == "Both sexes") |>
  select(iso3_code = spatial_dim_value_code, 
         year = period, 
         value = fact_value_numeric) |>
  tidy_df("alcohol_consumption_rate", average = TRUE)
```

### Outcomes of Violence
```{r import_violence_outcomes}
homicide_df = 
  read_excel(path = "data/unodc/intentional_homicide.xlsx", skip = 2) |>
  janitor::clean_names() |>
  filter(indicator == "Victims of intentional homicide",
         unit_of_measurement == "Rate per 100,000 population",
         dimension == "Total", 
         sex == "Total", 
         age == "Total") |>
  tidy_df("homicide_rate", average = TRUE)
  
violence_df = 
  read_excel(path = "data/unodc/violent_sexual_crime.xlsx", skip = 2) |>
  janitor::clean_names() |>
  filter(indicator == "Violent offences",
         unit_of_measurement == "Rate per 100,000 population") |>
  tidy_df("violence_rate", average = TRUE)
```

The datasets for economic crimes, criminal justice personnel, homicide and 
violence datasets provided data for both counts and rates per 100,000 people. 
Therefore, we filtered to choose rates per 100,000 people since it allows 
standardization for comparability by adjusting for differences in population 
size. 

### Merge Datasets
```{r merge_datasets}
merged_violence_df =
  left_join(homicide_df, violence_df) |>
  left_join(gdp_df) |>
  left_join(inflation_df) |>
  left_join(unemployment_df) |>
  left_join(human_develop_df) |>
  left_join(econ_crime_df) |>
  left_join(personnel_df) |>
  left_join(trafficking_df) |>
  left_join(alcohol_df) |>
  mutate(country = as.factor(country),
         region = as.factor(region)) |>
  filter(between(year, 2015, 2023)) |>
  relocate(iso3_code)

head(merged_violence_df)
```

The final merged dataset includes `r nrow(merged_violence_df)` rows and 
`r ncol(merged_violence_df)` columns, including iso3 codes, country, region, 
year, homicide rate, average violence offence rate, gdp, inflation rate, 
unemployment rate, average crime rate, criminal justice personnel rate, total, 
human trafficking rate and alcohol consumption rate as variables. There is a 
total of `r length(unique(pull(merged_violence_df, country)))` distinct 
countries. 

After merging the datasets, country and region were converted to categorical 
variables and the dataset was filtered for the years between 2015 and 2023. 

# EDA

## Visualization
```{r}
homicide_visual_df = 
    readxl::read_excel(
    path = "data/unodc/intentional_homicide.xlsx",
    skip = 2
  ) |>
  janitor::clean_names() |>
  filter(indicator == "Victims of intentional homicide",
         unit_of_measurement == "Rate per 100,000 population",
         category != "Total",
         sex != "Total",
         between(year, 2015, 2023)) |>
  select(country, region, category, sex, year, homicide_rate = value)
```

```{r}
homicide_visual_df |>
  group_by(region, year) |>
  summarize(avg_homicide_rate = mean(homicide_rate)) |>
  ggplot(aes(y = avg_homicide_rate, x = as.factor(year), fill = region)) +
  geom_col(position = "dodge", bin = 3.0) + 
  labs(x = "Year",
       y = "Average Homicide Rate",
       Title = "Average Homicide Rate Across Region")
```
```{r}
homicide_visual_df |>
  group_by(region, sex, year) |>
  summarize(avg_homicide_rate = mean(homicide_rate)) |>
  ggplot(aes(y = avg_homicide_rate, x = year, color = sex)) +
  geom_smooth(se = FALSE) +
  facet_grid(~region) + 
  labs(x = "Year",
       y = "Average Homicide Rate",
       Title = "Average Homicide Rate Trend")
```

## Data Transformation 

## Regression 






