---
title: "**P8105 Final Project: Determinants of Global Violence**"
author: "My An Huynh, Jeffrey Lin, Soo Min You, Hyun Kim, Malika Top"
output: html_document
---

```{r setup, include=FALSE}
library(readxl)
library(countrycode)
library(plotly)
library(tidyverse)

```

# *Motivation*

Globally, violence undermines social and economic development, with its impacts 
being more pronounced in "developing" countries. Communities lacking adequate 
resources or experiencing systemic inequities are more vulnerable to various 
health challenges, including violence. Factors that influence health also play 
a role in shaping violence, which in turn affects determinants such as housing, 
education, transportation, and economic conditions. This project seeks to 
explore the factors driving serious assault, sexual violence, and intentional 
homicide worldwide from 2015 to 2023. 

Selected this time frame because it includes significant global events, 
including the COVID-19 pandemic and various conflicts, which have greatly 
influenced patterns of violence. Also, even though there are a lot of missing 
data for year 2023, we decided to include the year because this time period 
provides insight into the contemporary factors shaping violence worldwide. 

# *Related Work*

Conflict trends in 2023 revealed increasing threats to the state of global peace. 
New York Times (NYT) Journalist, Marc Lacey, titled NYT’s annual photojournalism 
review [“A Weary World”](https://www.nytimes.com/interactive/2023/world/year-in-pictures.html). 
Though 2024 comes to a close , the conflicts in Russia, Gaza, Ethiopia, and 
Sudan rage on, and more understanding of violence trends is needed. Tensions 
within and across countries have major implications for not just conflict 
prevention but also social and economic preservation which are often more fragile 
in vulnerable areas.

# *Initial Questions*

1. How can we best predict violence and homicide rates (i.e. what are its indicators)? 
2. How do rates of indicators compare across global regions and countries? 
3. How do rates of violence and homicide compare across global and regions and countries?
4. What categories of violent are most prevalent? 

# *Data*

## Source

Since there are many determinants and indicators of violence, we chose the 
indicators and outcomes that seem most relevant and interesting in 
exploring violence from the following sources:

* United Nations Development Program (UNDP): 
  - Human Development Index
* United Nations Office of Drugs and Crime (UNODC)
  - Corruption and Economic Crime
  - Criminal Justice Personnel
  - Human Trafficking
  - Intentional Homicide
  - Violent and Sexual Crimes
* World Bank
  - Gross Domestic Product (GDP)
  - Inflation Rate (Measured by Consumer Price Index)
  - Unemployment Rate
* World Health Organization (WHO)
  - Alcohol Consumption

Intentional homicide, and violent and sexual crimes were chosen as the outcome 
variable to quantify violence. 

* Intentional homicide:
  - counting unit: number of homicide victims per 100,000 population
  - classification:
    - situational context: organized crime, interpersonal (excluding 
    familial/intimate), socio-political
    - relationship to perpetrator: intimate partner, family member, friend, 
    colleague, etc.
    - mechanism: firearm, weapon, physical force

* Violent and sexual crimes:
  - counting unit: number of individual offences per 100,000 population
  - classification of offenses:
    - rape
    - serious assault
    - kidnapping
    - sexual violence
  - NOTE: some countries used other counting unit (a series of offenses to form 
  a case, or multiple cases to form an investigation)

## Scraping Method

The datasets were downloaded from the official websites of the sources above. 
The names of the files were also changed accordingly for clarity and to avoid 
confusion. For example, the alcohol consumption data file was renamed from 
"data.csv" to "alcohol_consumption.csv".

## EDA Data

For EDA, since we are interested in exploring the trends of global violence, 
we imported and tidied the datasets for our two outcome variables: homicide 
and violence. 

### Homicide
```{r}
homicide_visual_df = 
    read_excel(path = "data/unodc/intentional_homicide.xlsx",
               skip = 2) |>
  janitor::clean_names() |>
  filter(unit_of_measurement == "Rate per 100,000 population",
         indicator == "Victims of intentional homicide",
         category != "Total",
         sex != "Total",
         between(year, 2015, 2023)) |>
  select(country, region, category, sex, year, 
         homicide_rate = value) |>
  mutate(category = str_replace_all(category, 
                                    "^Intimate partner or family member.*", 
                                    "Intimate partner or family member"),
         category = str_replace_all(category, 
                                    "^Socio-political homicide.*", 
                                    "Socio-political homicide"))

```

The tidied homicide dataset includes country, region, category of homicide 
victims, sex, year and victim rate of homicide. The dataset included data for 
both counts and rates per 100,000 people. Therefore, we filtered to choose rates 
per 100,000 people since it allows standardization for comparability by 
adjusting for differences in population size.

### Violence
```{r}
violence_visual_df = 
    read_excel(path = "data/unodc/violent_sexual_crime.xlsx",
               skip = 2) |>
  janitor::clean_names() |>
  filter(unit_of_measurement == "Rate per 100,000 population",
         between(year, 2015, 2023)) |>
  select(country, region, indicator, category, sex, year, 
         violence_rate = value) |>
  mutate(category = str_replace_all(category, 
                                    "^Sexual violence.*", 
                                    "Sexual violence"))
```

The tidied violence dataset includes country, region, indicator, category of 
violence, sex, year and violence rate. The indicator includes different rates 
for violence (violent offences and victims) which are explored further in 
EDA. Rates per 100,000 people were chosen also for this dataset for 
comparability. 

## Regression Data

For regression, the datasets for all the chosen predictors and outcomes are 
tidied and merged for modeling process and analysis. 

### Function to Tidy Datasets
```{r pivot_tidy_function}
tidy_df = function(data, variable, pivot = FALSE, average = FALSE) {
  
  if(pivot) {
    data = data |>
      pivot_longer(cols = -iso3_code,
                   names_to = "year",
                   values_to = variable) |>
      janitor::clean_names()
  }
  
  data = data |>
    mutate(year = as.numeric(year),
           iso3_code = str_replace_all(iso3_code, "^GBR.*", "GBR"),
           iso3_code = str_replace_all(iso3_code, "^IRQ.*", "IRQ"),
           country = countrycode(iso3_code, origin = "iso3c",
                                 destination = "country.name",
                                 nomatch = NA, warn = FALSE)) |> 
    drop_na(country)
  
  if(average) {
    
    if("region" %in% colnames(data)) {
      data = data |>
        group_by( country, region, year) |>
        summarize(!!variable := mean(value))
    }
    
    else {
        data = data |>
          group_by(country, year) |>
          summarize(!!variable := mean(value))
    }
    
  }
  
  return(data)
  
}
```

We defined a function `tidy_df()`to pipe after importing each dataset to 
automatically tidy the datasets. The `countrycode()` function standardizes the 
country names of each dataset since some of the country names are represented 
inconsistently across dataset files. For example, South Korea was represented 
as "Korea (Republic of)" and "Korea, Rep." 

Moreover, for the countrycode() function, NA is returned if no matches are 
found and warnings are also suppressed by `nomatch = NA` and `warn = FALSE` 
respectively since some datasets include iso3 codes for entities that are not 
countries. For example, EUU represents the European Union; however, the 
`countrycode()` function only recognizes iso3 codes for countries and returns a 
warning, but we are interested in countries and regions. A boolean parameter 
called `pivot` was included to indicate if a dataset needs to be pivoted and 
subsequently perform `pivot_longer()`, since some datasets had years organized 
as different columns, where each column represented values for that year. 

Also, some of the datasets include rates for different categories. For example, 
the data for economic crime includes crime rates for fraud, burglary and more. 
Therefore, another boolean parameter called `average` was included to pass an 
option to group by and summarize rates across categories and genders of 
different datasets, in order to provide a standardized measure and reduce 
complexity for analysis.

### Economic Determinants
```{r tidy_econ_determinants, message = FALSE}
gdp_df = 
  read_excel(path = "data/worldbank/gdp.xls",
             sheet = "Data", skip = 3, na = "") |>
  select(iso3_code = 2, "2015":"2023") |>
  tidy_df("gdp", pivot = TRUE)

inflation_df = 
  read_excel(path = "data/worldbank/inflation_rate.xls", 
             sheet = "Data", skip = 3, na = "") |>
  select(iso3_code = 2, "2015":"2023") |>
  tidy_df("inflation_rate", pivot = TRUE)

unemployment_df =
  read_excel(path = "data/worldbank/unemployment_rate.xlsx",
             range = "A1:BP437", na = "..") |>
  rename_with(str_replace_all, pattern = " \\[.*", replacement = "") |>
  select(iso3_code = 4, "2013":"2023") |>
  tidy_df("unemployment_rate", pivot = TRUE)

human_develop_df =
  read_csv(file = "data/undp/human_development_index.csv", na = "") |>
  head(-11) |>
  rename_with(str_replace_all, pattern = "hdi_", replacement = "") |>
  select(iso3_code = 1, "2015":"2022") |>
  tidy_df("hdi", pivot = TRUE)
```

The following variables were identified as economic prdictors of violence: 

* `gdp`
* `inflation`
* `unemployment`
* `human development index`

The `pivot` parameter of the `tidy_df()` function was set to true for these 
datasets, since all of them had the different years as data columns, which 
needs to be pivoted accordingly. 

### Social Determinants
```{r import_social_determinants, message = FALSE}
econ_crime_df = 
  read_excel(path = "data/unodc/corruption_economic_crime.xlsx", skip = 2) |>
  janitor::clean_names() |>
  filter(unit_of_measurement == "Rate per 100,000 population") |>
  tidy_df("crime_rate",  average = TRUE)

personnel_df =
  read_excel(path = "data/unodc/criminal_justice_personnel.xlsx", skip = 2) |>
  janitor::clean_names() |>
  filter(unit_of_measurement == "Rate per 100,000 population",
         indicator == "Criminal Justice Personnel", 
         sex == "Total") |>
  tidy_df("personnel_rate", average = TRUE)

trafficking_df =
  read_excel(path = "data/unodc/human_trafficking.xlsx", skip = 2) |>
  janitor::clean_names() |>
  filter(indicator == "Detected trafficking victims",
         category == "Total", 
         sex == "Total",
         age == "Total",
         txt_value != "<5") |>
  mutate(value = str_replace_all(txt_value, ",", ""),
         value = as.numeric(value)) |>
  select(iso3_code : region, year, trafficked_victims = value) |>
  tidy_df("trafficked_victims")

alcohol_df =
  read_csv(file = "data/who/alcohol_consumption.csv", na = "") |>
  janitor::clean_names() |>
  filter(dim1 == "Both sexes") |>
  select(iso3_code = spatial_dim_value_code, 
         year = period, 
         value = fact_value_numeric) |>
  tidy_df("alcohol_consumption_rate", average = TRUE)
```

The following variables were identified as social prdictors of violence: 

* `economic crime`
* `criminal justice personnel`
* `human trafficking`
* `alcohol`

The datasets for economic crimes and criminal justice personnel also provided 
data for both counts and rates per 100,000 people. Subsequently, we filtered to 
choose rates per 100,000 people for standardization and comparability. The 
human trafficking dataset provided data for number of human trafficking victims, 
instead of trafficking rates; the dataset also included a category for total 
number of trafficking victims, so the data was filtered for total category and 
neither the `pivot` nor `average` options were set to true in the `tidy_df()` 
function.

### Outcomes of Violence
```{r import_violence_outcomes, message = FALSE}
homicide_df = 
  read_excel(path = "data/unodc/intentional_homicide.xlsx", skip = 2) |>
  janitor::clean_names() |>
  filter(indicator == "Victims of intentional homicide",
         unit_of_measurement == "Rate per 100,000 population",
         dimension == "Total", 
         sex == "Total", 
         age == "Total") |>
  tidy_df("homicide_rate", average = TRUE)
  
violence_df = 
  read_excel(path = "data/unodc/violent_sexual_crime.xlsx", skip = 2) |>
  janitor::clean_names() |>
  filter(indicator == "Violent offences",
         unit_of_measurement == "Rate per 100,000 population") |>
  tidy_df("violence_rate", average = TRUE)
```

Unlike the tidied dataset for EDA, the datasets for homicide and violence had 
their respective rates averaged across category, since we are interested in 
the average rates of homicide victims and violent offences for regression 
analysis. 

### Merge Datasets
```{r merge_datasets, message = FALSE}
merged_violence_df =
  left_join(homicide_df, violence_df) |>
  left_join(gdp_df) |>
  left_join(inflation_df) |>
  left_join(unemployment_df) |>
  left_join(human_develop_df) |>
  left_join(econ_crime_df) |>
  left_join(personnel_df) |>
  left_join(trafficking_df) |>
  left_join(alcohol_df) |>
  mutate(country = as.factor(country),
         region = as.factor(region)) |>
  filter(between(year, 2015, 2023)) |>
  relocate(iso3_code) |>
  drop_na(violence_rate, homicide_rate)

head(merged_violence_df)
write.csv(merged_violence_df, file = "./data/merged_violence.csv")
```

The final merged dataset includes `r nrow(merged_violence_df)` rows and 
`r ncol(merged_violence_df)` columns, including iso3 codes, country, region, 
year, homicide rate, average violence offence rate, gdp, inflation rate, 
unemployment rate, average crime rate, criminal justice personnel rate, total, 
human trafficking rate and alcohol consumption rate as variables. There is a 
total of `r length(unique(pull(merged_violence_df, country)))` distinct 
countries. 

After merging the datasets, country and region were converted to categorical 
variables and the dataset was filtered for the years between 2015 and 2023, 
the period of our interest for analysis. Although the datasets were merged with 
matching country names tidied by the `countrycode()`, the iso3 codes were 
retained, since they are required for plotting maps which can be found under the 
shiny section of this website. 

# *Exploratory Data Analysis*

## Homicide Victims

### Global Trends of Homicide Victims
```{r message = FALSE}
homicide_visual_df |>
  group_by(region, sex, year) |>
  summarize(avg_homicide_rate = mean(homicide_rate)) |>
  ggplot(aes(y = avg_homicide_rate, x = year, color = sex)) +
  geom_smooth(se = FALSE) +
  facet_grid(~region) + 
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  labs(x = "Year",
       y = "Average Rate of Homicide Victims",
       title = "Homicide Victims Worldwide")
```

The Americas and Asia have significantly higher average homicide rate across 
years for male victims compared to the other regions. The average homicide rate 
between genders are not significant for Africa, Asia and Oceania. 

### Comparing the Americas and Asia
```{r message = FALSE, warning = FALSE}
homicide_visual_df |>
  filter(region %in% c("Asia", "Americas")) |>
  group_by(category, region) |>
  summarize(avg_homicide_rate = mean(homicide_rate)) |>
  ggplot(aes(y = avg_homicide_rate, x = region,
             fill = reorder(category, avg_homicide_rate))) +
  geom_col(position = "dodge", bin = 3.0) + 
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  labs(x = "Region",
       y = "Average Rate of Homicide Victims",
       title = "Homicide in America & Asia", 
       fill = "Category")
```

Further exploring homicide in Americas and Asia, the Americas have higher 
homicide rates for most categories compared to Asia. Homicides of perpetrator 
to victim relationship unknown has the largest homicide rate. Homicides of 
unknown types is next category with the highest homicide rate in the Americas. 

On the other hand, Asia has a significantly high homicide rate in 
socio-political homicide compared to the other categories,suggesting high 
casualties as a result of socio-political activities for countries in Asia. 
The other categories have a homicide rate lesser than 5 percent. 

### Breakdown of Homicide
```{r message = FALSE}
homicide_visual_df |>
  group_by(category, year) |>
  summarize(avg_homicide_rate = mean(homicide_rate)) |>
  ggplot(aes(y = avg_homicide_rate, x = as.factor(year), fill = 
               reorder(category, avg_homicide_rate))) +
  geom_bar(stat = "identity") + 
  labs(title = "Distribution of Homicide by Category",
       x = "Year",
       y = "Total Rate of Homicide Victims",
       fill = "Category")
```

Homicides of unknown types and perpetrator to victim relationship unknown are 
two highest proportions in the total homicide rate, indicating most homicides 
are committed by strangers or have no connection with the victims. Also, we can 
observe that homicide rates as a result of organized criminal groups or gangs 
and interpersonal homicide are the next significant proportions in homicide. 
The total homicide rate increases sharply from 2015 to 2017 compared to the 
other years possibly due to socioeconomic factors, such as economic instability. 


## Violence: Offences & Victims

### Global Trends of Violent Offences
```{r message = FALSE, fig.height = 10}
violence_visual_df |> 
  filter(indicator == "Violent offences") |> 
  group_by(category, year, region) |> 
  summarise(grp_mean_rate = mean(violence_rate)) |> 
  ggplot(aes(x = year, y = grp_mean_rate, color = category)) +
  geom_line() +
  facet_wrap(~region, ncol=2) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1),
        legend.position = "bottom") + 
  labs(title = "Violent Offences Worldwide",
       x = "Year", 
       y = "Average Rate of Violent Offences") +
  guides(color = guide_legend(nrow = 3))
```

Africa's most common violent offence category was serious assault, with the 
Americas trading off between serious assault and acts intended to induce fear 
or emotional distress. From 2015-2021 Oceania clearly has the highest rates of 
serious assault, but it drops severely in 2022. 

### Exploring Violent Offences
```{r message = FALSE, fig.height = 10, fig.width = 8}
violence_visual_df |>
  filter(indicator == "Violent offences") |>
  group_by(category, year, region) |> 
  summarise(avg_violence_rate = mean(violence_rate)) |> 
  mutate(category = fct_reorder(category, avg_violence_rate)) |>
  ggplot(aes(x = category, y = avg_violence_rate, fill = region)) +
  geom_bar(stat = "identity", position = "dodge") +
  facet_wrap(~year, ncol=2) +
  theme(axis.text.x = element_text(angle = 80, hjust = 1),
        legend.position = "bottom") +
  labs(x = "Category", 
       y = "Average Rate of Violent Offences",
       title = "Violent Offences by Category Across Time") 
```

The plot above shows similar trends to the line graph above, but allows us to compare more closely
the rates for categories that are less visible such as sexual violence, where Oceania appears
to have the highest rates of as well. We see more clearly that with Oceania's drop in serious
assaults in 2022, Americas and Oceania have similar trends.  
 

### Victims of Sexual Violence
Even in 2024, it is still ever-important to discuss sexual violence crimes. 
Violence against women is a public health issue and any type of sexual violence 
is a violation of human rights, harming physical, mental, sexual, and reproductive 
health. We are interested in seeing more specifically trends in sexual violence 
across the years for regions with sufficient data. 

```{r message = FALSE}
violence_visual_df |>
  filter(indicator == "Victims of sexual violence",
         region != "Africa") |>
  group_by(region, sex, year) |>
  summarize(avg_victim_rate = mean(violence_rate)) |>
  ggplot(aes(y = avg_victim_rate, x = year, color = sex)) +
  geom_smooth(se = FALSE) +
  facet_grid(~region) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  labs(x = "Year",
       y = "Average Rate of Sexual Violence Victims",
       title = "Sexual Violence Worldwide")
```

When exploring victim rate of sexual violence, we can see that in overall, 
there is a higher average victim rate for females; the largest disparity in 
victim rate between the genders can be observed in Oceania, where the rate for 
females have been sharply increasing over the years. The victim rate for the 
Americas and Europe has also been increasing, but steadily compared to Oceania 
over the years, while Asia has experienced a decline from 2019. Africa was not 
included because the data was either very small or missing for many countries 
in the region. 

### Investigating Sexual Violence
```{r message = FALSE}
violence_visual_df |>
  filter(indicator == "Victims of sexual violence",
         region != "Africa") |>
  group_by(category, sex, region) |>
  summarize(avg_victim_rate = mean(violence_rate)) |>
  ggplot(aes(y = avg_victim_rate, x = region, 
             fill = reorder(category, avg_victim_rate))) +
  geom_col(position = "dodge") + 
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  labs(x = "Region",
       y = "Average Rate of Sexual Violence Victims",
       fill = "Category",
       title = "Sexual Violence By Category Across Region")
```

Further exploring victims of sexual violence, except for perpetrator unknown to 
the victim, Oceania has a higher average victim rate of sexual violence for the 
all other categories compared to the other regions. Oceania, Europe and Americas 
have relationship of perpetrator is not known as the highest category for 
victim rate. Similar to homicide, the perpetrators are strangers or have no 
connection with the victims for most sexual violence. On the other hand, Asia 
has a significantly low average victim rate, having other perpetrator known to 
the victim as the category with the highest rate.

### Breakdown of Sexual Violence
```{r message = FALSE}
violence_visual_df |>
  filter(indicator == "Victims of sexual violence",
         region != "Africa") |>
  group_by(category, year) |>
  summarize(avg_victim_rate = mean(violence_rate)) |>
  ggplot(aes(y = avg_victim_rate, x = as.factor(year), 
             fill = reorder(category, avg_victim_rate))) +
  geom_bar(stat = "identity") + 
  labs(x = "Year",
       y = "Total Rate of Sexual Violence Victims",
       title = "Distribution of Sexual Violence by Category",
       fill = "Category")
```

Relationship to perpetrator is not known in the category with a significant 
proportion to total victim rate of sexual violence compared to the other 
categories across the years. The proportion of victim rates for the other three 
categories become approximately equal to each other from 2019. Specifically, 
the proportion for intimate partner or family member and perpetrator unknown to 
the victim increases over time.

## Global Prevalence of Violence

Across the time period from 2015 to 2023, the Americas consistently exhibited 
a significantly higher homicide rate compared to other regions. The difference 
in this region is pronounced, indicating a persistent and widespread issue with 
violent crime. The homicide rate in Asia is also notable. There was a rising 
trend in homicide rate from 2015 to 2021, but was followed by a sharp decrease 
in 2022. However, it should be noted that not all countries that reported 
homicide rate also reported the violence rate. 

The drop in 2022 may imply that there was a potential shift or improvement in 
the region, but it could also mean that there are some missing data in that 
year, leading a bias or misinterpretation. When examining violence rates, 
the Americas followed a similar trend to the homicide rate, with rates significantly 
higher than the other regions. This trend reflects the persistent issue of violence 
in the region, affecting not only homicide rates but also broader forms of violence. 
However, Oceania presented an extreme outlier in terms of victim rates of sexual 
violence, surpassing even the Americas.

Putting these results against our own knowledge, Oceania's rates seem higher
than expected. This raises the question of if higher rates mean more violent
countries or countries who just have reported more crimes. It is very important
to address that these outcome data are compiled by  UNODC but the data itself comes
from each nation's respective administrative registers, often managed by law enforcement
agencies. Additionally, UNODC comes by this data not through requirement but by the 
United Nations Survey of Crime Trends and Operations of Criminal Justice Systems (UN-CTS). 
This means that countries voluntarily supply their information. However, a safety-guard 
is that UNODC does check for consistency and coherence with other data sources. 


# *Regression*


(NEED TO ADD REGRESSION)


# *Citation*

* UNODC (2024), UNODC Research - Data Portal – Access and
functioning of justice. https://dataunodc.un.org/dp-cjs-persons-convicted (Accessed on Nov. 2024).
* UNODC (2024), UNODC Research - Data Portal – Corruption and
Economic Crime. https://dataunodc.un.org/dp-crime-corruption-offences (Accessed on Nov. 2024).
* UNODC (2024), UNODC Research - Data Portal - Violent and Sexual
Crime. https://dataunodc.un.org/dp-crime-violent-offences (Accessed on Nov. 2024)
