---
title: "Regression"
author: "My An Huynh"
date: "2024-11-15"
output: html_document
---

```{r setup, include=FALSE}
library(tidyverse)
library(readxl)
library(countrycode)
library(MASS)
library(glmnet)
library(GGally)
library(leaps)
set.seed(1)
```

# Helpful Functions
```{r}
best_subset = function(predictor, outcome, criterion) {
  
  optimal_subset <- 
    leaps(x = predictor, y = outcome, nbest = 3, 
      method = criterion, names = names(predictor))
  
  if (criterion == "Cp") {
    optimal_rsq <- optimal_subset[[criterion]] %>% min()
    optimal_subset_idx <- optimal_subset[[criterion]] %>% which.min()
  } else {
    optimal_rsq <- optimal_subset[[criterion]] %>% max()
    optimal_subset_idx <- optimal_subset[[criterion]] %>% which.max()
  }
  
  print(optimal_rsq) 
  print(optimal_subset$which[optimal_subset_idx,])

}
```

This function serves as a wrapper around the leaps function. It performs 
best subset variable selection and then prints out the optimal model diagnostic 
and the predictor variables included in the regression model. 

## Regression 

Split data into training and testing + visualize distributions 
```{r train/test + distribution}
train_df = sample_frac(merged_violence_df, size = 0.8)
test_df = anti_join(merged_violence_df, train_df)

plot_distributions = function(column, name) {
  
  if(is.numeric(column) & name != "year") {
    ggplot(train_df, aes(x = column)) +
    geom_density() +
    labs(title = paste("Distribution of", name))
  }
  
}

train_list = colnames(train_df)
map(train_list, \(x) plot_distributions(pull(train_df, x), x))
```

After looking at the distributions of all variables, all variables are skewed 
right, except for human development index which is bimodal and slightly 
left-skewed. I will apply ln transformations and box-cox transformations to 
these variables, and use the Shapiro-Wilk test to test for normality. Even 
though it is not necessary to normalize the predictors, this step will stabilize 
variance and reduce heteroscedasticity. I think it will be helpful for further 
steps with model.

Transformation step involves writing a function for ln transformation and a 
function for box_cox transformation. The functions will be mapped into the 
nested listcol which includes all continuous variables in the train_df dataset. 

```{r ln_transform}
ln_transform = function(value) {
  return(log(abs(value)))
}

ln_train_df = 
  train_df |>
  mutate(across(c(homicide_rate:alcohol_consumption_rate), 
                ln_transform))

map(train_list, \(x) plot_distributions(pull(ln_train_df, x), x))
```

```{r boxcox_transform, eval = FALSE, include = FALSE}
boxcox_transform = function(value) {
  value = value + abs(min(value, na.rm = TRUE)) + 0.00001
  
  boxcox_result = boxcox(value ~ 1, plotit = FALSE)
  lambda = boxcox_result$x[which.max(boxcox_result$y)]
  return((value^lambda - 1) / lambda)
}

boxcox_train_df = 
  train_df |>
  mutate(across(c(homicide_rate:alcohol_consumption_rate), 
                boxcox_transform))

map(train_list, \(x) plot_distributions(pull(boxcox_train_df, x), x))
```

# Statistical Learning Approaches

## Data Pre-processing 

```{r}
lambda = 10^(seq(-2, 2.75, 0.1))

homicide_df =
  merged_violence_df |> 
  dplyr::ungroup() |> 
  dplyr::select(
    homicide_rate, everything(), -violence_rate, -year, -country, -region, 
    -iso3_code) |> 
  drop_na()



```


## Lasso 
Use lasso to find the most important variables in predicting homicide rate and violence rate. 

### Predictors for violence rate:
To prepare data for lasso (fitting violence rate), I removed variables that are not predictors including iso3_code, country, region, year and homicide rate. I then dropped NA values before performing lasso. 

```{r violence}
lasso_violence = 
  merged_violence_df |> 
  dplyr::select(-iso3_code, -country, -region, -year, -homicide_rate) |> 
  drop_na()

x = model.matrix(violence_rate ~. - country - region, data = lasso_violence)[, -1]
y = lasso_violence |> pull(violence_rate)

lambda = 10^(seq(-2, 2.75, 0.1))

lasso_violence_fit = 
  glmnet(x, y, lambda = lambda)

lasso_violence_cv = 
  cv.glmnet(x, y, lambda = lambda)

lambda_violence_opt = 
  lasso_violence_cv[["lambda.min"]]

lasso_violence_fit = 
  glmnet(x, y, lambda = lambda_violence_opt)

lasso_violence_fit |> 
  broom::tidy()
```

The optimal lambda for violence is 0.794. GDP has the least influence on the model, followed by `trafficked_victims`. These covariates will be removed from the model. 


### Predictors for homicide rate. 
```{r homicide}
lasso_homicide = 
  merged_violence_df |> 
  dplyr::select(-iso3_code, -country, -region, -year, -violence_rate) |> 
  drop_na()

x = model.matrix(homicide_rate ~. - country - region, data = lasso_homicide)[, -1]
y = lasso_homicide |> pull(homicide_rate)

lambda = 10^(seq(-2, 2.75, 0.1))

lasso_homicide_fit = 
  glmnet(x, y, lambda = lambda)

lasso_homicide_cv = 
  cv.glmnet(x, y, lambda = lambda)

lambda_homicide_opt = 
  lasso_homicide_cv[["lambda.min"]]

lasso_homicide_fit = 
  glmnet(x, y, lambda = lambda_homicide_opt)

lasso_homicide_fit |> 
  broom::tidy()

```

The optimal lambda for homicide is 0.694.

# Ridge Regression 

```{r}
lambda = 10^(seq(-2, 2.75, 0.1))

lasso_fit =
  glmnet(x, y, lambda = lambda)

lasso_cv =
  cv.glmnet(x, y, lambda = lambda)

lambda_opt = lasso_cv[["lambda.min"]]

```


# Baseline MLR with additive effects from all possible predictor variables 

## Predict Homicide Rates
```{r}
baseline_MLR_hom <- lm(homicide_rate ~ gdp + inflation_rate + unemployment_rate +
                   hdi + crime_rate + personnel_rate + trafficked_victims + 
                    alcohol_consumption_rate , data = merged_violence_df)
summary(baseline_MLR_hom)
```
## Check Baseline Homicide Prediction Model for Colinearity

```{r}
vif_baseline_hom_MLR <- car::vif(baseline_MLR_hom)
vif_baseline_hom_MLR %>% 
  tibble(
    variable = names(vif_baseline_hom_MLR),
    VIF = vif_baseline_hom_MLR
  ) %>% 
  dplyr::select(variable, VIF) %>% 
  knitr::kable()

```



## Predict Violence Rates
```{r}
baseline_MLR_viol <- lm(violence_rate ~ gdp + inflation_rate + unemployment_rate +
                   hdi + crime_rate + personnel_rate + trafficked_victims + 
                    alcohol_consumption_rate , data = merged_violence_df)
summary(baseline_MLR_viol)

```
## Calculate VIF for baseline violence MLR
```{r}
vif_baseline_viol_MLR <- car::vif(baseline_MLR_viol)
vif_baseline_viol_MLR %>% 
  tibble(
    variable = names(vif_baseline_viol_MLR),
    VIF = vif_baseline_viol_MLR
  ) %>% 
  dplyr::select(variable, VIF) %>% 
  knitr::kable()
```


## Heat map and MLR 
I will use cor() to find the correlation between the predictors of homicide rate and violence. 
```{r}

cor_matrix = 
  cor(violence_predictors_df[, c("inflation_rate", "unemployment_rate", "hdi", "alcohol_consumption_rate")], use = "pairwise.complete.obs")


ggcorrplot(cor_matrix, 
           method = "circle",  
           type = "lower",  
           lab = TRUE,        
           lab_size = 3,      
           colors = c("blue", "white", "red"), # Color scale (blue = negative, red = positive)
           title = "Correlation Heatmap"
)

```
I found that there is moderate correlation between `hdi` and `alcohol_consumption_rate`, corr = 0.5. I will fit 2 MLR models, one with and one without the interaction between these two covariates. 


Fit MLR for homicide rate 
```{r}
hom_linear_model = lm(homicide_rate ~ inflation_rate + unemployment_rate + hdi + alcohol_consumption_rate, data = merged_violence_df) 
summary(hom_linear_model)

hom_interaction_model = lm(homicide_rate ~ inflation_rate + unemployment_rate + alcohol_consumption_rate * hdi, data = merged_violence_df) 

AIC(hom_interaction_model)
AIC(hom_linear_model)

vif_hom_linear_model <- car::vif(hom_linear_model)
vif_hom_linear_model %>% 
  tibble(
    variable = names(vif_hom_linear_model),
    VIF = vif_hom_linear_model
  ) %>% 
  dplyr::select(variable, VIF) %>% 
  knitr::kable()

```

The adjusted R-squared for the linear model is 0.08781 and its AIC is 5015.024. The adjusted R-squared for the interaction model is 0.106429 and its AIC is 5002.571. This means that the interaction model explains the variation in homicide rate better than the linear model. 

Fit MLR for violence rate 
```{r}
violence_linear_model = lm(violence_rate ~ inflation_rate + unemployment_rate + hdi + alcohol_consumption_rate, data = merged_violence_df) 
summary(violence_linear_model)

violence_interaction_model = lm(violence_rate ~ inflation_rate + unemployment_rate + alcohol_consumption_rate * hdi, data = merged_violence_df)
summary(violence_interaction_model)

AIC(violence_interaction_model)
AIC(violence_linear_model)

vif_violence_linear_model <- car::vif(violence_linear_model) 
vif_violence_linear_model %>% 
  tibble(
    variable = names(vif_violence_linear_model),
    VIF = vif_violence_linear_model
  ) %>% 
  dplyr::select(variable, VIF) %>% 
  knitr::kable()

```

## Criterion-based procedure

Select Data
```{r}
violence_df =
  merged_violence_df |> 
  dplyr::ungroup() |> 
  dplyr::select(
    violence_rate, everything(), -year, -homicide_rate, -country, -region, 
    -iso3_code) |>
  drop_na() 

violence_predictors <- violence_df %>% 
  dplyr::select(-violence_rate)

violence_outcome <- violence_df %>% 
  pull(violence_rate)

```


Perform regsubset to predict violence_rate using r-squared as criterion
```{r}
best_subset(predictor = violence_predictors, outcome = violence_outcome, 
            criterion = "adjr2")
```
Perform regsubset to predict violence_rate using Cp as criterion

```{r}
best_subset(predictor = violence_predictors, outcome = violence_outcome, 
            criterion = "Cp")
```

Select Data
```{r}
homicide_df =
  merged_violence_df |> 
  dplyr::ungroup() |> 
  dplyr::select(
    homicide_rate, everything(), -violence_rate, -year, -country, -region, 
    -iso3_code) |> 
  drop_na()

homicide_predictors <- homicide_df %>% 
  dplyr::select(-homicide_rate)

homicide_outcome <- homicide_df %>% 
  pull(homicide_rate)

```

Predict Homicide Rates using R-squared as criterion
```{r}
best_subset(predictor = homicide_predictors, outcome = homicide_outcome, 
            criterion = "adjr2")
```

Predict Homicide Rates using Cp as criterion
```{r}
best_subset(predictor = homicide_predictors, outcome = homicide_outcome, 
            criterion = "Cp")
```
Check for co-linearity in best subset violence model 
```{r}
subset_violence_lm <- lm(violence_rate ~ gdp + inflation_rate + 
                           crime_rate + personnel_rate + trafficked_victims +
                           alcohol_consumption_rate, data = violence_df)
vif_subset_violence_model <- car::vif(subset_violence_lm) 

vif_subset_violence_model %>% 
  tibble(
    variable = names(vif_subset_violence_model),
    VIF = vif_subset_violence_model
  ) %>% 
  dplyr::select(variable, VIF) %>% 
  knitr::kable()
  
  
```

Check for co-linearity in best subset homicide model
```{r}
subset_homicide_lm <- lm(homicide_rate ~ inflation_rate + unemployment_rate +
                           hdi + crime_rate, data = homicide_df)

vif_subset_homicide_model <- car::vif(subset_homicide_lm)
vif_subset_homicide_model %>% 
  tibble(
    variable = names(vif_subset_homicide_model),
    VIF = vif_subset_homicide_model
  ) %>% 
  dplyr::select(variable, VIF) %>% 
  knitr::kable()
```



The results from criterion-based procedures suggest that significant predictors for violence_rate are gdp, inflation_rate, crime_rate, personnel_rate, trafficked_victims and alcohol_consumption rate, with a total of 6 predictors. This yields the most appropriate Cp value (5.363), which is approximately close to the number of predictors, and highest adjusted R-squared (0.510). 

For homicide_rate, the results are not as straightforward. The best model based on Cp and R-squared seems to be the model with 4 predictors because it has the lowest Cp (6.36), and a decently high adjusted R-squared (0.331), indicating a good balance between fit and complexity. As more predictors are added, R-squared adjusts slightly but Cp levels off at around 6 predictors. The model with 4 predictors, which includes inflation_rate, unemployment_rate, hdi, personnel_rate, seems to be a better trade-off.


# Model Comparison

## Cross-Validate
```{r}

```


## Calculate RMSE 
