---
title: "Regression"
author: "My An Huynh"
date: "2024-11-15"
output: html_document
---

```{r setup, include=FALSE}
library(tidyverse)
library(readxl)
library(countrycode)
library(MASS)
```


## Regression 

Split data into training and testing + visualize distributions 
```{r train/test + visualization}
train_df = sample_frac(merged_violence_df, size = 0.8)
test_df = anti_join(merged_violence_df, train_df, by = "gdp")

plot_distributions = function(column) {
  
  data = pull(train_df, column)
  
  if(is.numeric(data)) {
    ggplot(train_df, aes_string(x = column)) +
    geom_density() +
    labs(title = paste("Distribution of ", column))
  }
}

train_list = colnames(train_df)
map(train_list, plot_distributions)

```

After looking at the distributions of all variables, all variables are skewed 
right, except for human development index which is bimodal and slightly 
left-skewed. I will apply ln transformations and box-cox transformations to 
these variables, and use the Shapiro-Wilk test to test for normality. Even 
though it is not necessary to normalize the predictors, this step will stablize 
variance and reduce heteroscedasticity. I think it will be helpful for further 
steps with model.

Transformation step involves writing a function for ln transformation and a 
function for box_cox transformation. The functions will be mapped into the 
nested listcol which includes all continuous variables in the train_df dataset. 

```{r transformations}
ln_transform = function(df) {
  df |> 
    mutate(across(everything(), ~ {
      if(any(!is.na(.) & . <= 0)) {
      . = . + abs(min(., na.rm = TRUE) + 0.0000000001)
    }
      log(.)
    }))
}

nested_train_df =
  train_df |> 
  nest(data = c(homicide_rate:alcohol_consumption)) |> 
  mutate(
    ln_transformed_data = map(data, ln_transform)
  ) |> 
  unnest(ln_transformed_data) |> 
  rename_with(~ paste0("ln_", .), where(is.numeric)) |> 
  rename(
    year = ln_year
  ) 

str(nested_train_df)
nested_train_df =
  nested_train_df |> 
  select(country, year, region, starts_with("ln"))

```

Working on box cox but the code doesn't work ....

```{r eval = FALSE}
box_cox_transform = function(df) {
  df |> 
    mutate(across(where(is.numeric), ~ {
      if (any(!is.na(.) & . <= 0)) {
        . <- . + min(., na.rm = TRUE) + 0.0000000001  
      }

      box_cox_result <- boxcox(. ~ 1, lambda = "all", plotit = FALSE)

      box_cox_tidy <- tidy(box_cox_result)
      
      optimal_lambda <- box_cox_tidy |> 
        filter(logLik == max(logLik)) |> 
        pull(lambda) 
      
      if (optimal_lambda == 0) {
        log(.) 
      } else {
        ((.)^optimal_lambda - 1) / optimal_lambda  
      }
    }))
}

```

