---
title: "Regression"
author: "My An Huynh"
date: "2024-11-15"
output: html_document
---

```{r setup, include=FALSE}
library(tidyverse)
library(readxl)
library(countrycode)
library(MASS)
```


## Regression 

Split data into training and testing + visualize distributions 
```{r train/test + visualization}
train_df = sample_frac(merged_violence_df, size = 0.8)
test_df = anti_join(merged_violence_df, train_df, by = "gdp")

plot_distributions = function(column, name) {
  
  if(is.numeric(column) & name != "year") {
    ggplot(train_df, aes(x = column)) +
    geom_density() +
    labs(title = paste("Distribution of", name))
  }
  
}

train_list = colnames(train_df)
map(train_list, \(x) plot_distributions(pull(train_df, x), x))
```

After looking at the distributions of all variables, all variables are skewed 
right, except for human development index which is bimodal and slightly 
left-skewed. I will apply ln transformations and box-cox transformations to 
these variables, and use the Shapiro-Wilk test to test for normality. Even 
though it is not necessary to normalize the predictors, this step will stablize 
variance and reduce heteroscedasticity. I think it will be helpful for further 
steps with model.

Transformation step involves writing a function for ln transformation and a 
function for box_cox transformation. The functions will be mapped into the 
nested listcol which includes all continuous variables in the train_df dataset. 

```{r transformations}
ln_transform = function(value) {
  return(log(abs(value)))
}

ln_train_df = 
  train_df |>
  mutate(across(c(homicide_rate:alcohol_consumption), 
                ln_transform))

map(train_list, \(x) plot_distributions(pull(ln_train_df, x), x))
```

Working on box cox but the code doesn't work ....

```{r eval = FALSE}
boxcox_transform = function(value) {
  value = value + abs(min(value, na.rm = TRUE)) + 0.00001
  
  boxcox_result = boxcox(value ~ 1, plotit = FALSE)
  lambda = boxcox_result$x[which.max(boxcox_result$y)]
  return((value^lambda - 1) / lambda)
}

boxcox_train_df = 
  train_df |>
  mutate(across(c(homicide_rate:alcohol_consumption), 
                boxcox_transform))

map(train_list, \(x) plot_distributions(pull(boxcox_train_df, x), x))
```

